---
title: Lab Ten -- Point Estimation # Title of the document

# Bibliography Details
bibliography: Lab10References.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab10WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
```

**Tips**

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.
* Make sure to use "enter" to make your code readable, so it doesn't run off the page. I switched the Quarto document to automatically render to .pdf, so you can see what you're submitting by default. You can switch back for the highlighting by moving the html block above the pdf block in the YAML header.
* Don't forget that you can copy-and-paste code when you're doing something similar as we did in class!
* Make sure to plot all computed probabilities. This is good `ggplot` practice AND will help make sure you don't make a mistake when computing the probability.


# Lab 8 Notes

##  Altham's Multiplicative Binomial Distribution

The PMF for Altham's Multiplicative Binomial Distribution is
$$f_X(x) = \frac{1}{c} {n \choose x} p^x (1-p)^{n-x} \theta^{x(n-x)}I(x \in \{0, 1, ..., n\})$$
where
$$c = \sum_{i=0}^n {n \choose i} p^i (1-p)^{n-i} \theta^{i(n-i)}.$$ 
Just so we start on the same page, I have included my `R` function for computing probability using this distribution.

```{r}
daltbinom <- function(x, size, prob, theta){
  normalizing.constant <- 0
  for(j in 0:size){
    normalizing.constant <- normalizing.constant + 
      choose(size, j) * prob^j * (1-prob)^(size-j) * theta^(j*(size-j))
  }
  # Probability Mass at x=x
  (1/normalizing.constant) *
    choose(size, x) * prob^x * (1-prob)^(size-x) * theta^(x*(size-x)) *
      (x>=0)*(x<=size)
}
```

# Altham's Multiplicative Beta Binomial Distribution
The PMF for Altham's Multiplicative Beta Binomial Distribution is
$$f_X(x) = \left[\int_{0}^1 \left(\frac{1}{c} {n \choose x} p^x (1-p)^{n-x} \theta^{x(n-x)}\right)\left(\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)} p^{\alpha-1}(1-p)^{\beta-1}\right) dp\right]I(x \in \{0, 1, ..., n\})$$

Just so we start on the same page, I have included my `R` function for computing probability using this distribution.
```{r}
daltbbinom.single <- function(x, size, alpha, beta, theta){
    integrand <- function(p) {
    sapply(p, # integrate will pass in a vector, so we use sapply to apply the following
           function(p_val){
             # f(x|p) * f(p|alpha, beta)
             daltbinom(x, size, p_val, theta) * dbeta(p_val, alpha, beta)
           }
    )
  }
  # Probability Mass at x=x
  # integrate(integrand, lower = 0, upper = 1)$value * (x>=0)*(x<=size)
  # I shrink the bounds below to avoid log(0)=-infinty
  # I use try() to catch when there is an error and avoid crashing
  res <- try(integrate(integrand, lower = 1e-7, upper = 1 - 1e-7)$value * 
                  (x>=0)*(x<=size), 
             silent = TRUE)
  # If there is an error, return a near-zero probability
  if(inherits(res, "try-error")) return(0.1^6)
  
  res
}
# Write a function that works on a vector
daltbbinom <- function(x, size, alpha, beta, theta){
  sapply(X=x, FUN=daltbbinom.single, size=size, alpha=alpha, beta=beta, theta=theta)
}
```

# Question 1

In Lab 8, we use Altham's Multiplicative Binomial Distribution. Your job was to interpret the impact of $\theta$ in that equation. In this lab, we will model the number of legs won in a best of eleven (first to six) match. 

To help us think about the interpretation, something that was challenging before, let's try something a little more manageable.

## Part a
Suppose a player has a 50\% chance of winning a specific leg, that is let $p=0.50$. Let's look at the probability they win ($x=6$) legs in the match. Compute this probability over a sequence from $\theta=0.5$ to $\theta=1.5$ and plot it using a line where the $x$-axis is $\theta$ and the $y$-axis is $P(X=6)$.

solution:
```{r}
#| size: scriptsize
library(tidyverse) 

# created sequence of thetas
thetas <- seq(from = 0.5, to = 1.5, by = 0.001)

# created probabilities using daltbinom function from 1.1
probabilities <- sapply(thetas, function(theta){
  daltbinom(x = 6, size = 11, prob = 0.50, theta = theta)
})

# created dataframe consisting of my thetas and probabilities
dataframe = data.frame(new_theta = thetas, prob = probabilities)

# created plot
ggplot(dataframe) +
  geom_line(aes(x = new_theta, y = prob)) +
  xlab("theta") +
  ylab("P(X = 6)") +
  theme_bw()
```

## Part b
Now, plot Altham's Multiplicative Binomial Distribution under the same parameters with $\theta=0.50$, $\theta=1$, and $\theta=1.5$. Use `ncol=1` in `facet_wrap()` to plot the PMFs in a single column for comparing.

solution:
```{r}
#| size: scriptsize
library(tidyverse) 
library(patchwork)

# created a function for making my plots
plot_function <- function(theta) {
  x_values <- 0:11


# created probabilities using daltbinom function from 1.1
  probabilities <- sapply(x_values, function(x){
    daltbinom(x = x, size = 11, prob = 0.50, theta = theta)
  })

# created dataframe consisting of my thetas and probabilities
  dataframe = data.frame(x = x_values, prob = probabilities)

# created plot
  ggplot(dataframe) +
    geom_linerange(aes(x = x, ymin = 0, ymax = prob)) +
    labs(title = paste("Theta =", theta),
    x = "Number of Wins",
    y = "Probability") +
    theme_bw()
}

# used patchwork to plot the three plots next to each other
plot_function(0.5) | plot_function(1) | plot_function(1.5)
```

## Part c
What is your best assessment of what $\theta$ is doing in this setting. Consider when a 6-0 blowout is most likely, and when a 5-6 nail-biter is most likely. Note here we should treat the match as a fixed $n=11$ as we did for $n=3$ in Lab 8.\
**Note:** In lab 8, we saw that small $\theta$ meant wins came in bunches (clumped together), so there were more 2-0 wins in a best of three series.

Solution:
A game where a 6-0 blowout is most likely is when the theta is equal to 0.5 as the probabilities are highest at 0 or 11 wins. A game where a 5-6 nail biter is most likely is probably when theta is equal to 1.5 as the probabilities are highest at 5 and 6 wins.

# Question 2

## Part a
Altham's Multiplicative Binomial Distribution does not typically simplify to the clean $E(X)=np$ seen in the Binomial distribution, unless $\theta = 1$. Write out the expression that would need to be solved to find $E(X^k)$, and then write a function that computes it given the proposed parameter values $p$ (`prob`) and $\theta$ (`theta`).

Solution: 
E(X^k) = sum from 0 to infinity of x^k * fx(x)
```{r}
#| size: scriptsize
# created function for observations portion
E_altbinom_k <- function(k, size, prob, theta){
  x_values <- 0:size

# created function for proportion of obs = x portion
  probabilities <- sapply(x_values, function(x){
    daltbinom(x = x, size = 11, prob = 0.50, theta = theta)
  })

# combined the two portions to get my E(x^k)
  sum((x_values^k) * probabilitiess)
}
```

## Part b
Describe how you can use your answer to part a to find Method of Moments estimates for the parameters $p$ and $\theta$ for Altham's Multiplicative Binomial Distribution based on data.

Solution:
To find method of moments estimates for the parameters, p and theta, I would pass data through the function I created above and compute the k sample moments. I would then find the expected values or population moments by using the function and then set the sample moments equal to the population moments to solve for the parameters. I would set this up for each power from x^1 to x^k and then solve the system of equations.

## Part c
Write a function that computes the log-likelihood for Altham's Multiplicative Binomial Distribution, given data (`x`) and proposed parameter values $p$ (`prob`) and $\theta$ (`theta`).

Solution:
```{r}
#| size: scriptsize
# created log likelihood function
loglikelihood_altbinom <- function(parameters, x, n){
  p <- parameters[1]
  theta <- parameters[2]

# created function for probabilities
  probabilities <- daltbinom(x, n, p, theta)

# summation to calculate log likelihood
  loglikelihood <- sum(log(probabilities))
  return(loglikelihood)
}
```

## Part d
Describe how you can use your answer to part c to find Maximum Likelihood estimates for the parameters $p$ and $\theta$ for Altham's Multiplicative Binomial Distribution based on data.

Solution:
To find the maximum likelihood estimates I would take data and plug it into my function to multiply the probabilities of each observation together ("and" probability). If many probabiltiies between 0 and 1 are getting multiplied together, the resulting probability will be smaller so we take the log likelihood because it is easier to see the results. The goal is to find the parameters that make the observed data most probable so we optimize/maximize to find parameters that yield the highest combined probability of all of the observations. 

# Lab 10
Peter "Snakebite" Wright is a two-time Professional Darts Corporation World Champion (2020, 2022) and a former World No. 1. He is one of the sport's most decorated players, winning nearly 50 PDC titles including the World Matchplay, UK Open, and multiple European Championships. 

In 2024, he made The Premier League -- a league involving the eight best players that runs every Thursday night from February to May. In this season, he won two of eighteen matches, securing only four points. The spread in legs won was -43, meaning he lost 43 more legs than he won. This performance was rather quite bad. He was soundly trounced in almost all his match ups. 


The 2024 standings are below:

| Rank | Player             |
|:-----|:-------------------|
| 1    |Luke Littler        |
| 2    |Luke Humphries      |
| 3    |Michael van Gerwen  |
| 4    |Michael Smith       |
| 5    |Nathan Aspinall     |
| 6    |Rob Cross           |
| 7    |Gerwyn Price        |
| 8    |Peter Wright        |


## Summarize the 2024 Data

In `pdc2024.csv`, I have provided the data for the 2024 Premier League season. There are a few data cleaning steps to consider.

  * Remove any matches with no `Score` (e.g., a bye or walkover)
  * Nights 1 through 16 are best of 11 (first to 6), but the playoffs are extended. For consistency, keep nights 1 through 16 only.
  * Make two columns `WinnerLegs` and `LoserLegs` that keep track of the number of legs each player has won
  
  Summarize the data. What do the data tell us about the breakdown of the results?

Solution:
```{r}
#| size: scriptsize
library(tidyverse)
pdc_data <- read.csv("pdc2024.csv")
pdc_cleaned <- pdc_data |>
  filter(Score != "w/o") |> # removed walkovers
  mutate(Night = as.numeric(Night)) |> # made night column all numbers instead of strings
  filter(Night <= 16) |> # kept only nights 1-16
  separate(Score, into = c("WinnerLegs", "LoserLegs"), sep = "-", convert = TRUE) 
  # created two new columns "WinnerLegs" and "LoserLegs" that keep track of the number of legs each player has won

# created summary statistics table
pdc_cleaned |>
  count(WinnerLegs, LoserLegs) |>
  mutate(propotion = n / sum(n))
```

The most common score was 6-5 with about 28% of games ending with that score followed closely by 6-4 and 6-3, meaning most games were competetive. Blowouts where pretty infrequent, at least in comparison to close games. 

## Altham's Multiplicative Binomial Distribution
For each player in the 2024 season, use the number of legs won in each match to compute maximium likelihood estimates for $p$ and $\theta$. \
**Note**: The "for each" here can be done more efficiently with a function or a loop!

* Create a vector containing the number of legs won by the player. Note sometimes you will need to pull from winner's legs and sometimes from the loser's legs.
* Find the MLE for the player using this data. Ensure to report $p$ and $\theta$ for each player in a nicely formatted table.\
**Note:** In Lecture on Monday, we used transformations to restrict how `optim()` searches the parameter space. For example, here, you might consider
`p <- 1/(1+exp(-par[1]))` and `theta <- exp(par[2])`.
* Make comments about the MLEs in the context of the standings.

Solution:
```{r}
#| size: scriptsize

# Created function to get the number of legs won by the player
player_legs <- function(data, player_name){
  new_data <- data |> 
    mutate(LegsWon = ifelse(Winner == player_name, WinnerLegs, LoserLegs)) |>
    filter(Winner == player_name | Loser == player_name)
  new_data$LegsWon
}

# created loglikelihood function:
loglikelihood <- function(par, x, n = 11, neg = FALSE){
  # restrict how optim() searches the parameter space (from lab) 
  p <- 1 / (1 + exp(-par[1])) # p between 0 and 1
  theta <- exp(par[2]) # theta positive
  ll <- sum(log(daltbinom(x, n, p, theta)))
  # Return negative if needed
  ifelse(neg, -ll, ll)
}

# created MLE function for one player 
MLE <- function(x){
  estimate <- optim(par=c(0,0), fn=loglikelihood, x=x, neg=TRUE) #same method from class notes
  p_player <- 1 / (1 + exp(-estimate$par[1])) # p between 0 and 1
  theta_player <- exp(estimate$par[2]) # theta positive
  return(c(p = p_player, theta = theta_player))
}

# created vector of all winners and losers and then used unique function to get rid of repeats
players <- unique(c(pdc_cleaned$Winner, pdc_cleaned$Loser))

#created initialized vectors
p_values <- numeric(length(players))
theta_values <- numeric(length(players))

# created for loop to get estimates for each player.
for(i in 1:length(players)){
  # used functions created above
  x <- player_legs(pdc_cleaned, players[i])
  estimate <- MLE(x)
  # updated the initialized vectors
  p_values[i] <- estimate["p"]
  theta_values[i] <- estimate["theta"]

# created dataframe with all the estimates
results <- data.frame(Player = players,
                      p = p_values,
                      theta = theta_values)
}

results
```

The players at the top of the 2024 standings shown above have the highest p values (Littler with 0.469 and Humphries with 0.461), meaning they win more and win a higher proportion of legs. A player like Peter Wright has the lowest p value and is last in the standings, meaning he wins fewer legs on average. Players like Luke Littler and Luke Humphries have higher thetas (1.316 and 1.205 respectively). This likely means that their games closer and more competetive (closer to 6-5 or 6-4) where as someone like Peter Wright has a lower theta (0.989) implying that his games are closer to blowouts, with him likely losing more often then not based off his p value and the standings


## Altham's Multiplicative Beta Binomial Distribution

For each player in the 2024 season, use the number of legs won in each match to compute maxmium likelihood estimates for $p$ and $\theta$.\
**Note**: The "for each" here can be done more efficiently with a function or a loop!

* Create a vector containing the number of legs won by the player. Note sometimes you will need to pull from winner's legs and sometimes from the loser's legs.
* Find the MLEs for the player using this data. Ensure to report $\alpha$, $\beta$, and $\theta$ for each player in a nicely formatted table.\
**Note 1**: Due to the computational complexity of `integrate()`, you should compute the logged probability for `0:6` once and add them according to the data in each player's vector.
**Note 2**: In Lecture on Monday, we used transformations to restrict how `optim()` searches the parameter space.
* Plot the underlying Beta($\alpha$, $\beta$) distribution for each player. \
**Note**: Add the mean and variance of the Beta to the table of MLEs. Recall
\begin{align*}
 E(X)   &= \frac{\alpha}{\alpha+\beta}\\
 sd(X) &= \sqrt{\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}}
\end{align*}
* Make comments about the MLEs in the context of the standings.\
**Note**: The players alternate who throws first in a leg has the mathematical "head start" to reach a finish before their opponent can take another turn. Players alternate who throws first with each leg.

```{r}
#| size: scriptsize

# Created function to get the number of legs won by the player (copied from 5.2)
player_legs <- function(data, player_name){
  new_data <- data |> 
    mutate(LegsWon = ifelse(Winner == player_name, WinnerLegs, LoserLegs)) |>
    filter(Winner == player_name | Loser == player_name)
  new_data$LegsWon
}

# created function to get log of probabilities
log_probs <- function(x_values, n, alpha, beta, theta){
  # created function to get normal probabilities
  probabilities <- sapply(x_values, function(x){
    daltbbinom(x, n, a = alpha, b = beta, theta = theta)
  })
  # took log of probabilities
  log(probabilities)
}

# created loglikelihood function:
bbloglikelihood <- function(par, x, n = 11, neg = FALSE){
  # restrict how optim() searches the parameter space (from lab). Make sure all params are positive (use exp) 
  alpha <- exp(par[1]) 
  beta <- exp(par[2])
  theta <- exp(par[3])
  x_values <- 0:6
  # used log probs function above to get log probs
  log_prob <- log_probs(x_values, n, alpha, beta, theta)
  # summed log probs like I did in 5.2 for each observation. Had to use x+1 to go through the x values 
  ll <- sum(log_prob[x+1])
  # Return negative if needed
  ifelse(neg, -ll, ll)
}

# created MLE function for one player 
bbMLE <- function(x){
  estimate <- optim(par=c(0,0,0), fn=bbloglikelihood, x=x, neg=TRUE) # same as 5.2 with one more parameter
  alpha_player <- exp(estimate$par[1]) # alpha must be positive
  beta_player <- exp(estimate$par[2]) # beta must be positive
  theta_player <- exp(estimate$par[3]) # theta must be positive
  # returned same as MLE function in 5.2
  return(c(alpha = alpha_player, beta = beta_player, theta = theta_player))
}

# created vector of all winners and losers and then used unique function to get rid of repeats
players <- unique(c(pdc_cleaned$Winner, pdc_cleaned$Loser))

#created initialized vectors
alpha_values <- numeric(length(players))
beta_values <- numeric(length(players))
theta_values <- numeric(length(players))

# created for loop to get estimates for each player.
for(i in 1:length(players)){
  # used functions created above
  x <- player_legs(pdc_cleaned, players[i])
  estimate <- bbMLE(x)
  # updated the initialized vectors
  alpha_values[i] <- estimate["alpha"]
  beta_values[i] <- estimate["beta"]
  theta_values[i] <- estimate["theta"]
}

# mean and sd of beta (directly from lab)
beta_mean <- alpha_values / (alpha_values + beta_values)
beta_sd <- sqrt((alpha_values*beta_values)/((alpha_values + beta_values)^2 * (alpha_values + beta_values + 1)))

# created dataframe with all the estimates (same as 5.2 with mean and sd now)
results <- data.frame(Player = players,
                      alpha = alpha_values,
                      beta = beta_values,
                      theta = theta_values,
                      mean = beta_mean,
                      sd = beta_sd)
results
```

```{r}
#| size: scriptsize
library(tidyverse)

# created initialized, empty vectors
x_vals <- seq(0, 1, by = 0.01)
x_values <- c()
y_values <- c()
player_values <- c()

# created for loop to loop over players. Updated y values using alpha and beta from above and updated initialized vectors
for(i in 1:length(players)){
  y_vals <- dbeta(x_vals, alpha_values[i], beta_values[i])
  x_values <- c(x_values, x_vals)
  y_values <- c(y_values, y_vals)
  player_values <- c(player_values, rep(players[i], length(x_vals)))
}

# created dataframe with data from above
df <- data.frame(x = x_values, y = y_values, Player = player_values)

# created plot
ggplot() +
  geom_line(data = df, aes(x = x, y = y)) +
  facet_wrap(~Player, ncol = 1, scales = "free_y") +
  # my plot came back looking very squished so I looked up how to change the scale and it told me to include "free_y"
  theme_bw()
```


## Executive Summary
Summarize the work you've done here to provide a brief (200-300 word) data-driven summary of the 2024 Premier League season using your work in Lab 10. Your summary should be professional, clear, and all claims should be corroborated with statistical support.


Solution:
The 2024 Premier League Darts season was able to be summarized by Altham's multiplicative binomial and beta binomial models. The estimated leg-winning probability (p in the binomial case and the beta means (alpha/(alpha+beta)) in the beta binomial case) correlated with how individuals players performed in the standings. Players towards the top of the standings like Like Littler and Like Humphries had both higher means (0.425 and 0.403 respectively) when compared with worse players like Peter Wright (0.197). The highest probabilities of winning individual legs directly coincided with winning more overall games and therefore placing higher in the standings. The standard deviation showed how consistent/variable the outcomes of specific players were. Players like Peter Wright and Nathan Aspinal had lower standard deviations (0.125 and 0.172 respectively), meaning they had more consistent outcomes. Players like Like Littler, Michael Van Gerwin and Luke Humphries had higher standard deviations (all around .36 or .38), implying they had more variability and showed their matches and outcomes were more unpredictable. Similarly to the Altham Multiplicative Binomial model, the theta in the beta-binomial model indicated how competitive or close matches typically were for certain players. For someone like Luke Littler, with a theta of 10.797, his games were typically more competitive than someone like Peter Wright who had a theta of 1.21. This likely implies that Luke Littler won a lot, but most of his games were very tight, while Peter Wright lost a lot and was getting blown out in almost all of his matches. In general, the models showed that higher ranked players win more legs on average and seem to have much closer and consistent performances when compared with those towards the bottom of the standings.