---
title: Unit 3 - Everything You Need to Know About Probability Distributions # Title of the document

# Bibliography Details
bibliography: ProbabilityReview.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab8WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
```

# Introduction to Probability
In data science, we rarely have the luxury of seeing the "whole picture." This is largely due to the fact that being able to attain data of entire country, population, or demographic can both be extremely time consuming and expensive. Due to this, we work with samples—small windows into a larger reality. The goal of probability is to take these samples and guess the **blueprint** of the entire population. 

To accomplish this we will follow this guideline: 

1. Establish What a Random Variable Is
2. Dive Into the Discrete Distribution
3. Dive Into the Continuous Distribution

# The Random Variable

The way to describe the **blueprint** is to use two ideas, namely the **random variable** and the **support**

The **random variable** is like a numerical bridge. It takes a real-world event (like a car crash) and turns it into a number ($X$) that math can understand. It is can be a characteristic, attribute, measurement, or any descriptive piece of data to depict an observation. 

Additionally, when talking about random variables, you have to mention the **Support**. The **Support** is the list of "allowable" realities.Analogy: If you are at a restaurant, the Support is the menu (everything you could order). The **Distribution** is the record of what people actually ordered. 

The random variable can take on different forms  either being classified as **quantitative** or **categorical**

1. A **Quantitative** variable implies that the data assumes numerical values. 
2. A **Categorical** variable implies the data are grouped (either by a common attribute, or category, or species, or any other grouping descriptor). Examples include: the species of an animal.

For our purposes, we will be looking specifically at **Quantitative** variables. Quantitative variables can take on two forms (or **Distributions**) amongst themselves. They are either:

1. **Discrete** - Implies the variable can be counted (e.g., 1, 2, 3, 4, or the number goals scored in a sporting game).
2. **Continuous** - Implies the variable can take on an infinite number of values in a given range. Continuous variables are measured and can represent any value along a continuum, such as time, weight, or distance as opposed to being measured in whole numbers.

# The Discrete Distribution

When our support (or "menu") consists of distinct, countable outcomes (like the number of goals in a match or relapses in a year) we use Discrete models. Discrete models use three different tools to illustrate how the data looks.

The first is called the **Probability Mass Function (PMF)**. This is our tool for precision. It answers the question: "What are the odds of seeing exactly $x$?". An example for this could be if you flipped a coin 10 times, what are the odds of landing on heads exactly six times. Mathematically, we define this as:

$$f_X(x) = P(X = x)$$

The second tool is called the **Cumulative Density Function (CDF)**. A lot of times, an analyst doesn't care about the probability of being "exactly 2." They care about risk: "What is the probability there are at most two goals scored in a game." This is the Cumulative Distribution Function (CDF). It acts as a running total of our PMF. Mathematically it can be defined as: 

$$F_X(x) = P(X \leq x) = \sum_{i \in S, i \leq x} f_X(i)$$

The final tool for the discrete distribution is the **Inverse CDF**. In the soccer example, the CDF tells you: "What is the probability of scoring 2 or fewer goals?" The Inverse CDF flips the script and shows: "How many goals do I need to score to be in the top 10% of teams?" The mathematical definition is:

$$F_X^{-1}(p) = \inf \{x \in \mathbb{R} : F_X(x) \geq p\}$$

To illustrate the discrete distribution and it's tools, I used data from the English Premiere League from [@EPL2024] to collect data on the number of goals scored per match. The data follows a Poisson Distribution with lambda = 1.39.

```{r}
#| size: scriptsize
# loaded data as epl_raw
epl_raw <- read_csv("E0.csv")

# Got number of goals and combined home and away into one 'population'
epl_goals <- epl_raw |>
  select(FTHG, FTAG) |>
  pivot_longer(cols = everything(), 
               names_to = "team_type", 
               values_to = "goals_scored")

# Calc avg goal rate (Lambda)
avg_goals <- mean(epl_goals$goals_scored)
```

```{r}
#| label: highlighted-soccer-plots
#| fig-cap: "Visualizing the probability of scoring exactly 1 goal (PMF) and at most 1 goal (CDF)"
#| fig-width: 5
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: scriptsize

# caluclated stored average goals as lambda
lambda.dat <- avg_goals
ggdat <- tibble(x = 0:10) |> 
  mutate(PMF = dpois(x, lambda = lambda.dat),
         CDF = ppois(x, lambda = lambda.dat))

# created PMF Plot 
PMF.plot <- ggplot() +
  geom_linerange(data = ggdat,
                 aes(x = x, ymin = 0, ymax = PMF)) +
  geom_linerange(data = ggdat |> filter(x == 1),
                 aes(x = x, ymin = 0, ymax = PMF),
                 color = "darkorchid2", linewidth = 2, alpha = 0.50) +
  geom_hline(yintercept = 0) +
  scale_x_continuous("Goals Scored (x)", breaks = 0:10) +
  ylab("Probability") +
  theme_bw()

# created the CDF Plot
CDF.plot <- ggplot() +
  geom_step(data = ggdat, aes(x = x, y = CDF)) +
  geom_point(data = ggdat, aes(x = x, y = CDF)) +
  geom_point(data = ggdat |> filter(x <= 1),
             aes(x = x, y = CDF),
             color = "darkorchid2", size = 2.5, alpha = 0.50) +
  geom_hline(yintercept = 0) +
  scale_x_continuous("Goals Scored (x)", breaks = 0:10) +
  ylab("Cumulative Probability") +
  theme_bw()

# plotted using patchwork
PMF.plot | CDF.plot
```

The PMF (left plot) shows the  probability of a game having exactly one goal scored in the game highlighted in purple. The mathematical equation to represent that can be seen below: $$f_X(1) = P(X = 1)$$

The CDF (right plot) shows the probability of a game having at most one goal scored. Notice how in the CDF, the height of the line is the summation of the individual probabilities illustrated in the left plot. This can be seen as the probability of zero goals being scored is about 0.23. The probability of one goal being scored is about 0.35. The probability of at most one goal being scored (one goal or zero goals) is equal to 0.58 which is a summation of the first two numbers. The mathematical equation can be seen below: $$F_X(1) = P(X \leq 1) = \sum_{i \in S, i \leq x} f_X(i)$$

# The Continuous Distribution

